A Bidirected String Graph Model for Genome Assembly

### 2012-12-10 presentation/paper.tex
paper: updates

Sequencing an organism's genome is an important step to further genetic research
on that organism.  However, currently available sequencing technology can only
sequence short fragments of DNA.  As a result, whole-genome shotgun sequencing
approaches to genome sequencing rely on programs called {\em genome assemblers}
that can put together a set of short DNA fragments to reconstruct the original
genome.  This paper provides an introduction and formalization of the genome
assembly program, then introduces a model and algorithm for genome assembly that
is based on a special type of graph called a \KeyTerm{bidirected string graph}.
\end{abstract}

@@

@@

Therefore, to sequence an entire genome, which may be millions of base pairs
long, the DNA from the genome must be broken into much smaller fragments, which
are then sequenced.  Following this, the original genome is computationally
reconstructed by a computer program called a \KeyTerm{genome assembler} in a
process called \KeyTerm{genome assembly}.  \KeyTerm{Genome sequencing} therefore
refers to the entire process of sequencing a genome, including both lab work and
computational work, while \KeyTerm{genome assembly} only refers to the
computational work.\footnote{It usually also takes some technology-specific
computational work to actually prepare the sequences for assembly; for example,
a series of digital photographs of clumps of fluorescently labeled molecules
might need to be interpreted as reads using image-processing algorithms.
However, these programs and/or algorithms generally have already been developed
by the manufacturer of the sequencing machine and work fairly well, and ``genome
assembly'' does not cover these algorithms, although genome assembly algorithms
still may be tuned for specific sequencing technologies.}

@@

It is also possible for reads to come from foreign DNA; for example, if a DNA
sample is obtained from grinding up plant or animal tissue, there could be some
bacterial cells, each with its own genome, contaminating the sample.  Therefore,
ideally an assembler should be prepared to completely ignore some reads.
However, this problem is not considered in this paper.

For the purposes of this paper, the problem of genome assembly is simplified by
assuming that the reads are error-free and uniformly (randomly) sampled from the
genome.  In addition, it is assumed there are no reads from foreign DNA, and the
reads are assumed to be unpaired, which means that mate-pair information, if
available, is not used.  In other words, a read is defined here as an exact
substring from either strand of any part of the genome.

Genomes may contain \KeyTerm{repeats}, which are subsequences that occur
multiple times in the genome--- possibly adjacent to each other (a
\KeyTerm{tandem repeat}), but not necessarily.  Repeats may be longer than the
length of any individual read, which makes the genome more difficult to assemble
accurately.  However, real genomes almost always contain repeats of substantial
length, so genome assembly algorithms must be able to handle repeats.

@@

Genome assembly is superficially similar to the shortest common superstring
(SCS) problem, which is to find the shortest possible string that contains every
string in a given set as substrings.  Genome assembly also requires finding a
longer string that contains all of a set of substrings.  However, the actual
genome may be more than one string, and it genome may be longer than the
shortest possible genome from which the reads could have been sampled, due to
repeats.  In addition,  genomes are double-stranded, and there may be other
complications with the reads (described in subsection \ref{subsec:reads}), such
as sequencing error.  Therefore, genome assembly is not the same as the SCS
problem, and in fact seems to be harder, despite the fact that the SCS problem
is already NP-complete\cite{Turner1989}.

In fact, some models of genome assembly have also been proven to be NP-complete
\cite{Medvedev2007}.  However, it is not acceptable to have a genome assembly
algorithm that takes exponential time to complete.  Instead, heuristic
algorithms that tend to produce good results but also tend to have a reasonable
running time must be designed.

@@

This paper explores the genome assembly problem through an algorithm published
in 2005 by Eugene Myers\cite{Myers2005} that is based on the idea of the
fragment string assembly graph.  The algorithm provides a good overview of the
techniques that are used in genome assembly, especially since it draws on ideas
used in previous assemblers, such as the Celera assembler, which was used to
assemble the human genome for the first time\cite{Venter2001}.  In addition, the
algorithm shares some ideas with de Bruijn graph-based assemblers, such as
EULER-SR\cite{Pevzner2001}, although it does not use a de Bruijn graph for its
assembly model.  The algorithm also introduces some new ideas.

The overall fragment string graph algorithm is as follows: first, use overlaps
between reads to build a graph that models the assembly problem.  The idea is
that walks through this graph model a way in which reads may be assembled
together when the overlaps are used as evidence for joining two reads.  The next
step is to simplify and analyze the graph; and finally, find paths through the
graph are found to reconstruct the original genome.

@@

			%\STATE{Maximally extend the occurrence of $\SeedVar$ at
			%$\OccOne$ and $\OccTwo$ outwards in both directions}
			%\STATE{${\tt pos1} \gets$ the beginning of the maximally
			%matched region on the read containing $\OccOne$}
			%\STATE{${\tt pos2} \gets$ the beginning of the maximally
			%matched region on the read containing $\OccTwo$}
			%\STATE{${\tt plen} \gets$ the length of the maximall
			%matched region on the read containing $\OccTwo$}
		\ENDFOR
	\ENDFOR
\end{algorithmic}
\end{algorithm}
\end{comment}

@@

A \KeyTerm{walk in a bidirected graph} $G$ is a continuous sequence of edges in
$G$ such that if any vertex $v$ is entered through a head inwards, it is exited
through a head outwards (unless it is the end of the path), and vice versa.  An
equivalent formulation is that if we consider a possible walk in a bidirected
graph, such as the possible walk $1 \to 2 \to 3 \to 4$ in the graph shown in
\Figure{fig:bidiwalk_example}, in order to check whether the walk is valid or
not, we must consider every vertex in the possible walk except for the very
first and last vertices.  For each such vertex, we need to verify that each time
that vertex is passed through as part of a walk, the arrow heads adjacent to
that vertex that are used in the relevant part of the walk are pointed in the
same direction {\em relative to the walk}. If there are no such pairs of heads
that point towards each other or point away from each other, the walk is valid.
The red circles in \Figure{fig:bidiwalk_example} show the two pairs of heads
that need to be checked for the possible walk $1 \to 2 \to 3 \to 4$.

@@

\begin{figure}
	\begin{center} {\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (2.00, 0) {2};
			\node (3) at (4.00, 0) {3};
			\node (4) at (6.00, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
			\draw[color=red,thick] (2.00, 0) circle (0.8);
			\draw[color=red,thick] (4.00, 0) circle (0.8);
		\end{tikzpicture}
	} \end{center}
	\caption{$1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid walks in
	the bidirected graph below.  (Check circled pairs of heads.)}
	\label{fig:bidiwalk_example}
\end{figure}

A \KeyTerm{cycle} in a bidirected graph is a walk that begins and ends at the
same vertex in such a way that it would be valid to trace the walk again when
the starting vertex is reached at the end of the cycle.

A bidirected graph is \KeyTerm{connected} if the undirected graph produced by
replacing all bidirected edges with undirected edges is connected.  A bidirected
graph is \KeyTerm{strongly connected} if it is connected and for every edge
there is a cycle containing it\cite{Medvedev2009}.

The concept of a walk through a bidirected graph maps quite nicely onto the
genome assembly problem, since a walk through the bidirected string graph
constructed from reads and overlaps models a way in which the reads can be
assembled together consistently.  In such a walk, each read may be used in
forward or reverse-complement orientation, depending on the direction, relative
to the walk, of the adjacent arrow heads when the corresponding vertex is passed
through.

@@

These interpretations can be generalized to the other two types of bidirected
edges.  Furthermore, these intepretations imply that the actual sequence spelled
out by a walk in the bidirected graph may include implied (non-labeled) sequence
at the start (if we start at the end of a read) and/or the end (if we end at the
beginning of a read).

@@

As \Figure{fig:E_coli} shows, the bidirected string graph may still be very
complicated even after transitive reduction and traversal.  In even worse
examples, the graph could easily have thousands of vertices and edges.  We still
need to figure out how to walk through the graph to produce the best assembly.

An \KeyTerm{Eulerian path in a bidirected graph} is a walk that traverses each
bidirected edge exactly one time (only in one of the two possible directions).
Suppose we could find an Eulerian path in the bidirected string graph.  Then
this path would correspond to an assembly that accounts for every overlap and
therefore would be a valid assembly of the reads.  Furthermore, such a path may
allow for vertices to be used for than one time in the assembly to allow for
genomic repeats, which would not be true if we found a Hamiltonian path
instead.\footnote{A Hamiltonian path is also inappropriate here because
unbranched paths have been collapsed into single edges, which {\em must} be
included in the assembly if the reads corresponding to the inner vertices are to
take part in the assembly.  Furthermore, keep in mind that there is not a
one-for-one correspondence between vertices and reads because contained reads
were removed (see \Section{sec:removing_contained_reads)}.}

However, there are multiple problems with the Eulerian path approach.  First,
the graph may not be Eulerian.  In the case that the graph

>>>>

Sequencing an organism's genome is an important step to further genetic research
on that organism.  However, currently available sequencing technology can only
sequence short fragments of DNA.  As a result, the whole-genome shotgun
sequencing approach to genome sequencing relies on breaking up a sample of
genomic DNA into fragments, sequencing those fragments, then using a program
called a {\em genome assembler} to put together these short sequences to
reconstruct the original, much longer sequence.  This paper provides an
introduction and formalization of the genome assembly problem, then introduces a
model and algorithm for genome assembly that is based on a special type of graph
called a \KeyTerm{bidirected string graph}.  \end{abstract}

@@

\subsection{Complement and reverse-complement sequences}

The \KeyTerm{complement} of a single-stranded DNA sequence $S$ is the
single-stranded DNA sequence obtained by making the replacements $\Base{A}
\leftrightarrow \Base{T}$ and $\Base{C} \leftrightarrow \Base{G}$.  The
\KeyTerm{reverse-complement} of a single-stranded DNA sequence $S$ is the
complement of $S$ reversed.  Physically, the reverse-complement of a
single-stranded DNA sequence corresponds to the corresponding sequence on the
other strand of the DNA molecule.  It is the reverse-complement, not the
``regular'' complement, because the two strands of a DNA molecule run in
opposite directions.

The \KeyTerm{canonical form} of a DNA sequence is the lexicographically lesser
of that sequence and its reverse-complement.  This concept is useful in
algorithms where a DNA sequence and its reverse-complement are considered
equivalent.

@@

Therefore, to sequence an entire genome, which may be millions of base pairs
long, the DNA from the whole genome must be broken into much smaller fragments,
which are then sequenced.\footnote{This specifically refers to the
\KeyTerm{whole-genome shotgun sequencing} approach to genome assembly rather
than alternatives such as \KeyTerm{BAC-to-BAC} sequencing that are able to
restrict sequences to certain regions of the genome in order to make sequencing
the genome easier (currently, at the disadvantage of increased cost and
non-computational work).} Following this, the original genome must be
computationally reconstructed by a computer program called a \KeyTerm{genome
assembler} in a process called \KeyTerm{genome assembly}.  \KeyTerm{Genome
sequencing} therefore refers to the entire process of sequencing a genome,
including both lab work and computational work, while \KeyTerm{genome assembly}
only refers to the computational work.\footnote{It usually also takes some
technology-specific computational work to actually prepare the sequences for
assembly; for example, a series of digital photographs of clumps of
fluorescently labeled molecules might need to be interpreted as reads using
image-processing algorithms.  However, these programs and/or algorithms
generally have already been developed by the manufacturer of the sequencing
machine and work fairly well, and ``genome assembly'' does not cover these
algorithms, although assemblers still may be tuned for specific sequencing
technologies.}

@@

It is also possible for reads to come from foreign DNA; for example, if a DNA
sample is obtained from grinding up plant or animal tissue, there could be some
bacterial cells, each with its own genome, contaminating the sample.  Ideally,
an assembler should be prepared to completely ignore a read if it seems too
unusual.  However, this problem is not considered in this paper.

For the purposes of this paper, the problem of genome assembly is simplified by
assuming the following characteristics of the reads:

\begin{itemize}
	\item The reads are error-free and uniformly sampled from the genome.
	\item There are no reads from foreign DNA.
	\item The reads are unpaired, meaning that mate-pair information, if
available, is not used.
\end{itemize}

In other words, a read is defined here as an exact substring from either strand
of any part of the genome.

Genomes may contain \KeyTerm{repeats}, which are subsequences that occur
multiple times in the genome--- possibly adjacent to each other (a
\KeyTerm{tandem repeat}), but not necessarily.  Repeats may be longer than the
length of any individual read, which makes the genome more difficult to assemble
accurately.  However, real genomes almost always contain repeats of substantial
length, so assemblers must be able to handle repeats.

@@

Genome assembly is superficially similar to the shortest common superstring
problem, which is to find the shortest possible string that contains every
string in a given set as substrings.  Genome assembly also requires finding a
longer string that, roughly, contains every string in  a set of substrings.
However, the actual genome may be more than one string, and the genome may also
be longer than the shortest possible genome from which the reads could have been
sampled, due to repeats.  In addition,  genomes are double-stranded, and there
may be other complications with the reads, such as sequencing error.  Therefore,
genome assembly is not the same as the shortest common superstring problem, and
in fact seems to be harder, despite the fact that the shortest common
superstring problem is already NP-complete\cite{Turner1989}.

In fact, some models of genome assembly have also been proven to be
NP-complete\cite{Medvedev2007}.  However, it is not acceptable to have a genome
assembly algorithm that takes exponential time to complete.  Instead, heuristic
algorithms that tend to produce good results on real inputs while still having
reasonable running times must be designed.

@@

This paper explores the genome assembly problem through an algorithm published
in 2005 by Eugene Myers that is based on the idea of the fragment string
assembly graph\cite{Myers2005}.  The algorithm provides a good overview of the
techniques that are used in genome assembly, especially since it draws on ideas
used in previous assemblers, such as the Celera assembler, which was used to
assemble the human genome for the first time\cite{Venter2001}.  In addition, the
algorithm shares some ideas with de Bruijn graph-based assemblers, such as
EULER-SR\cite{Pevzner2001}, although it does not use a de Bruijn graph for its
assembly model.  In fact, the algorithm is called the {\em fragment} string
graph algorithm because the graph is built directly from the sequenced fragments
(or reads), rather than subsequences ($k$-mers) taken from them.

The algorithm also introduces new ideas that were not used in previous
publications or assemblers.

The overall fragment string graph algorithm can be described as follows.  First,
use overlaps between reads to build a graph that models the assembly problem.
The idea is that walks through this graph model a way in which reads may be
assembled together consistently when the overlaps are used as evidence for
joining pairs of reads.  The next step is to simplify and analyze the graph, and
the last step is to find paths through the graph to reconstruct the original
genome.\footnote{The fragment string graph algorithm, as described in
\cite{Myers2005}, does not involve any finishing steps such as scaffolding
(using mate-pair information to improve the assembly), although ideas for doing
this are outlined in \cite{Medvedev2009}.  However, in this paper we assume that
no mate-pair information is available.}

@@

			%\STATE{Maximally extend the occurrence of $\SeedVar$ at
			%$\OccOne$ and $\OccTwo$ outwards in both directions}
			%\STATE{${\tt pos1} \gets$ the beginning of the maximally
			%matched region on the read containing $\OccOne$}
			%\STATE{${\tt pos2} \gets$ the beginning of the maximally
			%matched region on the read containing $\OccTwo$}
			%\STATE{${\tt plen} \gets$ the length of the maximally
			%matched region on the read containing $\OccTwo$}
		\ENDFOR
	\ENDFOR
\end{algorithmic}
\end{algorithm}
\end{comment}

@@

The \KeyTerm{in-degree} of a vertex $v$ in a bidirected graph is the number of
heads adjacent to $v$ that point inwards, while the \KeyTerm{out-degree} of $v$
is the number of heads adjacent to $v$ that point outwards.  A bidirected graph
$G$ is \KeyTerm{balanced} if for every vertex in $G$ the in-degree is equal to
the out-degree.

A \KeyTerm{walk} in a bidirected graph $G$ is a continuous sequence of edges in
$G$ such that if any vertex $v$ is entered through a head inwards, it is exited
through a head outwards (unless it is the end of the path), and vice versa.  An
equivalent formulation is that if we consider a possible walk in a bidirected
graph, such as the possible walk $1 \to 2 \to 3 \to 4$ in the graph shown in
\Figure{fig:bidiwalk_example}, in order to check whether the walk is valid or
not, we must consider every vertex in the possible walk except for the very
first and last vertices.  For each such vertex, we need to verify that each time
that vertex is passed through as part of a walk, the arrow heads adjacent to
that vertex that are used in the relevant part of the walk are pointed in the
same direction {\em relative to the walk}. If there are no such pairs of heads
that point towards each other or point away from each other, the walk is valid.
The red circles in \Figure{fig:bidiwalk_example} show the two pairs of heads
that need to be checked for the possible walk $1 \to 2 \to 3 \to 4$.

@@

\begin{figure}
	\begin{center} {\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (2.00, 0) {2};
			\node (3) at (4.00, 0) {3};
			\node (4) at (6.00, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
			\draw[color=red,thick] (2.00, 0) circle (0.8);
			\draw[color=red,thick] (4.00, 0) circle (0.8);
		\end{tikzpicture}
	} \end{center}
	\caption{$1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid
	walks in the bidirected graph above.  (Check circled pairs of heads to
	verify this.)}
	\label{fig:bidiwalk_example}
\end{figure}

A \KeyTerm{cycle} in a bidirected graph is a walk that begins and ends at the
same vertex in such a way that it would be a valid to trace the walk again when
the starting vertex is reached at the end of the cycle.

A bidirected graph is \KeyTerm{connected} if and only if  the undirected graph
produced by replacing all bidirected edges with undirected edges is connected as
defined for undirected graphs.  A bidirected graph is \KeyTerm{strongly
connected} if and only if it is connected and for every edge there is a cycle
containing it.\cite{Medvedev2009}

The concept of a walk in a bidirected graph maps quite nicely onto the genome
assembly problem because a walk through the bidirected string graph constructed
from a set of reads and their overlaps models a way in which the reads can be
assembled together consistently.  In such a walk, each read may be used in
forward or reverse-complement orientation, depending on the direction, relative
to the walk, of the adjacent arrow heads when the corresponding vertex is passed
through.

@@

These interpretations can be generalized to the other two types of bidirected
edges.  Furthermore, these interpretations imply that the actual sequence spelled
out by a walk in the bidirected graph may include implied (non-labeled) sequence
at the start (if we start at the end of a read) and/or the end (if we end at the
beginning of a read).

@@

As \Figure{fig:E_coli} shows, the bidirected string graph may still be very
complicated even after transitive reduction and traversal.  In even worse
examples, the graph could easily have thousands of vertices and edges.
Therefore, we need to figure out how to walk through the graph to produce the
best assembly.

An \KeyTerm{Eulerian path in a bidirected graph} is a walk that traverses each
bidirected edge exactly one time in either direction.  Suppose we could find an
Eulerian path in the bidirected string graph.  Then this path would correspond
to an assembly that accounts for every overlap and therefore would be a valid
assembly of the original reads.  Furthermore, such a path may allow for vertices
to be used for than one time in the assembly to allow for genomic repeats, which
would not be true if we found a Hamiltonian path instead.\footnote{A Hamiltonian
path is also inappropriate here because unbranched paths have been collapsed
into single edges, which {\em must} be included in the assembly if the reads
corresponding to the inner vertices are to take part in the assembly.
Furthermore, keep in mind that there is not a one-for-one correspondence between
vertices and reads because contained reads were removed (see
\Subsection{subsec:removing_contained_reads}).}

However, there are multiple problems with the Eulerian path approach.  First,
the graph may not be Eulerian.  This can happen if the graph is not strongly
connected, not balanced (indegree always equal to out
